Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-03-09 ms.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.  % change 32
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. Additionally, we investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.  Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  
Each year, approximately 10 million people die from cancer globally \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance. %change 30
One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis. Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}. %change 4
Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation. % change x1

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires an adaptation that increases fitness in the new environment such that extinction is averted.
There are three potential ways for a population to survive environmental change \citep{alexander2014evolutionary,bell2017evolutionary}: % change 45 %change 28
migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; 
adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. 
Here, we focus on the latter.

There has been extensive theoretical analysis of evolutionary rescue.
\citet{orr2008population,orr2014population} thoroughly investigated the simplest case in which there is a specific single mutation or low-frequency variant that is sufficient to rescue the population. 
\citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue;
however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced.
%\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
and found that for some parameter values rescue by two mutational steps is more likely than rescue by a single step.

Resistance to cancer therapy is a natural application of evolutionary rescue theory \citep{alexander2014evolutionary}. 
\citet{iwasa2003evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction, focusing on the case in which the cancer must get two mutations to be rescued, with the single-mutant still dying rapidly.
\citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
A very similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}.
\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells with one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
The division and death rate difference is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.

We assume that both chromosomal missegregation and mutations occur during the process of mitosis. 
Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_s$.
Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{s}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment. We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift.

@@

The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
\[
\begin{aligned}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{aligned}
\]

@@

Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug. Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common). We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics.

All the parameters discussed above are shown in \Cref{table1}.

@@

In our model, \emph{evolutionary rescue} occurs when drug-resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($s_t=a_t=m_t=0$). %change r1
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population. Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection. In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress. % change 53 
Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely \citep{iwasa2003evolutionary}, and the rescue probability is given by 
\begin{equation}\label{eq:rescue_prob} 
\presc = 
1-\left(1-p_s\right)^N \approx
1-\e^{-Np_s} = 
1-e^{-N/N^*} ,
\end{equation}
where the approximation $(1-p_s)\approx e^{-p_s}$ assumes that $p_s$ (but not necessarily $N p_s$) is small.
Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$, and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.

We use multi-type branching-process theory to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in three distinct scenarios (Appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
In these approximations, an important quantity is $T^* = \sqrt{\lambda_m / 4 v \lambda_a^2 \Delta_m}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} ,
\end{equation}
which is similar to a classical result by \citet{orr2008population}. %change 33
Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it resuced the threshold tumor size for rescue. 
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

In our analysis, we used branching processes, which assume that growth (division and death) is density-independent. However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
Therefore, we performed stochastic simulations of a logistic growth model with a carrying capacity. 
We find that our density-independent approximations agree with the results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).

@@

In the above, we assumed that at the onset of drug treatment, the initial tumor consisted entirely of drug-sensitive cells.
However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}. Moreover, aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}.
Hence, if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).

Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{c}{\abs{\Delta_s}}.
\end{equation}
This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.  
In contrast, \textit{de novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs. \textit{de novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \textit{de novo} genetic variation is  the main driver of evolutionary rescue.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.  % change 32
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 %change S2
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. We also investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue. %change S2
 Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  
Each year, approximately 10 million people die from cancer globally \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance. %change 30
One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
Aneuploidy is caused by chromosomal instability and missegregation of chromosomes during mitosis. % change S4
Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}. %change 4
Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.  A cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation. %change S16
Some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation. % change x1 %change S16

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires an adaptation that increases fitness in the new environment such that extinction is averted.
There are three potential ways for a population to survive environmental change \citep{alexander2014evolutionary,bell2017evolutionary}: % change 45 %change 28
migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; 
adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
and adaptation through genetic modifications, like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
Here, we focus on the latter.

There has been extensive theoretical analysis of evolutionary rescue.
\citet{orr2008population,orr2014population} thoroughly investigated the simplest case in which there is a specific single mutation or low-frequency variant that is sufficient to rescue the population. 
\citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number. 
Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue; however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced. %change S19
%\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
and found that for some parameter values rescue by two mutational steps is more likely than rescue by a single step.

Resistance to cancer therapy is a natural application of evolutionary rescue theory \citep{alexander2014evolutionary}. 
\citet{iwasa2003evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction, focusing on the case in which the cancer must get two mutations to be rescued, with the single-mutant still dying rapidly.
\citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
A similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}. % change S1
\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells with one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
The division and death rate difference is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[tolerant cells are able to survive the effects of the drug for an extended period of time compared to susceptible cells while resistant cells require a higher drug concentration to be affected, see][for a detailed explanation of the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. % change S1 %change S17

We assume that both chromosomal missegregation and mutations occur during the process of mitosis. 
Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_s$.
Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{s}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment. We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of aneuploids and selection against them, without the effect of genetic drift. %change RS3

@@

The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
\[
\begin{aligned}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)&&\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t&&\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t&&\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t&&\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)&&\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t&&\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t&&\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t&&\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t&&\left(\text{death of mutant cell}\right).
\end{aligned}
\]% change S6

@@

Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (see \Cref{table1}). %change S7
Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common). We assume deleterious mutations and other aneuploidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics. %change RS4

@@

In our model, \emph{evolutionary rescue} occurs when drug-resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($s_t=a_t=m_t=0$). %change r1
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population. Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection. In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress. % change 53 
Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is likely \citep{iwasa2003evolutionary}, and the rescue probability is given by % change S1
\begin{equation}\label{eq:rescue_prob} 
\presc = 
1-\left(1-p_s\right)^N \approx
1-\e^{-Np_s} = 
1-e^{-N/N^*} ,
\end{equation}
where the approximation $(1-p_s)\approx e^{-p_s}$ assumes that $p_s$ (but not necessarily $N p_s$) is small.
Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$, and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.

We use multi-type branching-process theory to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in three distinct scenarios (Appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
In these approximations, an important quantity is $T^* = \sqrt{\lambda_m / 4 v \lambda_a^2 \Delta_m}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is sufficiently rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that % change S1
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} ,
\end{equation}
which is similar to a classical result by \citet{orr2008population}. %change 33
Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size neeede for rescue. %change S8
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

In our analysis, we used branching processes, which assume that growth (division and death) is density-independent. However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
 We performed stochastic simulations of a logistic growth model with a carrying capacity. %change S16
We find that our density-independent approximations agree with the results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).

@@

We assumed that at the onset of drug treatment the initial tumor consisted entirely of drug-sensitive cells but aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}. %change S9, S19 
 Aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
If the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment). %change S16, S19

Therefore, the threshold tumor size for rescue by standing genetic variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that %change S10
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{c}{\abs{\Delta_s}}.
\end{equation}
This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.  
In contrast, \textit{de novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$). % change S1
For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs. \textit{de novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \textit{de novo} genetic variation is  the main driver of evolutionary rescue.
